### 2022-09-27 AISTATS2023/paper.tex
updated method description

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one the ALE approximation. Regarding the
definition, i.e., the formulation of the feature effect, ALE does not
inform the user about the level of heterogeneous effects hidden behind
the average effect (global explanation). In contrast, in the case of
PDP, the heterogeneous effects can be (partially) spotted by exploring
the Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
e.g.~the estimation of ALE from the limited samples of the training
set, ALE requires an important additional step, called
\textit{bin-splitting}. Bin-splitting consists of partitioning the
axis of the feature of interest in a sequence of non-overlapping
intervals and estimating a single effect from the population of
samples in each interval. Specifying an approriate sequence of
intervals is of particular importance, since ALE's interpretation is
meaningful only inside each interval \citep{molnar2022}. However, this
crucial step has not raised the appropriate attention, making the
approximation vulnerable to potential misinterpretations.

@@

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). The
black-box function is \(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and
the FE of the \(s\)-th feature is \(f^{\mathtt{<method>}}(x_s)\),
where \(\mathtt{<method>}\) is the name of the FE method.\footnote{An
  extensive list of all symbols used in the paper is provided in the
  helping material.}

\subsection{Feature Effect Methods And ALE Definition.}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|X_s = x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as \(f^s(z, \xc) = \dfdx (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
averaged local effects:

@@

FE methods pose the question \textit{what is expected to happen to the
  output (expected effect), if the value of a specific feature is
  increased/decreased}. Having quantified the expected effect, it
comes naturally to also ask \textit{how certain we are about the
  expected change (uncertainty)}. For this reason, the quantification
of the uncertainty, whether there are local explanations that deviate
from the expected global effect, has attracted a lot of interest.

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are produce a plot plot on top-of PDP. Both methods produce one curve
for each instance of the training set, i.e. ICE the instance effect
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) and d-ICE the derivative
effect, i.e.
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\). The user can visually examine the level of homogeneity of
the set of curves as an indicator of the level of heterogeneous
effects. Some methods try to automate the aforementioned visual
exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
  herbinger2022repid}. Some other approaches, like H-Statistic,
Greenweal, quantify the level of interaction between the input
features. A strong interaction index is an indirect indicator for the
existence of heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features, as we will analyze later. To the best
of our knowledge, no method so far quantifies the heterogeneous
effects on-top of ALE.

@@

In real ML scenarios, the FE is estimated from the limited instances
of the training set. Therefore, \citep{apley2020visualizing} proposed
dividing the \(s\)-th axis in \(K\) bins and estimating the local
effects in each bin by evaluating the black box-function at the bin
limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE)
approximation for computing the local effects on the training
instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \frac{\partial f}{\partial x_s}(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins \(K\) for splitting the axis into K equally-sized bins, an
approach with crucial limitations. Firstly, it is vulnerable to poor
bin-effect approximations that lead to a misleading ALE plot. On the
one hand, setting a \(K\) to a small value may hide fine-grain effects
due to large bins and setting \(K\) to a high value leads to
estimations of bin-effect from very limited samples. In general, the
user may face contradictory explanations for differnt \(K\) without a
clue to decide which one to trust. Secondly, as indicated
by~\citep{molnar2022}, in ALE the interpretation of the effect can
only be local. Preseting a fixed-size intervals, does not permit the
extraction of local regions (intervals) ng of data-driven local
regions

@@

The NAME method extends ALE for quantifying the level of heterogeneous
effects (uncertainty) and automating the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
presenting the component that extends ALE with uncertainty
quantification. In Section~\ref{sec:interval-based-estimation}, we
provide an alternative definition of NAME based on variable-size bins,
providing important remarks and proofs for connecting it with the
initial definition. In Section~\ref{sec:interval-spliting}, we define
the problem of optimal bin-splitting, i.e. how to optimally
approximate METHOD based on limited samples, and we provide an
algorithmic solution to the problem. Finally, in
Section~\ref{sec:visualization}, we illustrate the appropriate
visualization of NAME for facilitating its interpretation by a
non-expert and we discuss important aspects of the method.

@@

ALE does not shed light to potential heterogeneous effects that
contribute to the average explanation. Therefore, we extend ALE
definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
NAME method formulates the effect at a specific point \(x_s\) with a
compact entity that consists of the average effect and the uncertainty
\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\). In Section~\ref{sec:visualization},
we propose an appropriate visualization for easier interpretation of
NAME.

@@

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). For this reason, we firstly define the mean
effect (bin-effect) and the uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\):

@@

%
We also define as \(\mathcal{Z}_s\) a sequence of \(K+1\) points that
partitions the \(s\)-th feature axis into a sequence \(K\)
variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine NAME using an interval-based formulation
\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}
%
The proof is in the Appendix.

@@

The two definitions of NAME are equivalent, i.e.,
\(f^{\mathtt{ALE}}_{\mu}(x_s) = \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)\)
and
\(f^{\mathtt{ALE}}_{\sigma}(x_s) =
\tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), if and only if the sequence
of limits \(\mathcal{Z}\) is such that
\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\), where \(\rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial z\).

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}\) of the dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}= \{ \mathbf{x}^i : z_1 \leq x^i_s < z_2 \} \), i.e.
\(\hat{\mu}(z_1, z_2) = \frac{1}{|\mathcal{S}|} \sum_{i:\mathbf{x}^i
  \in \mathcal{S}} \left [ \dfdx(\mathbf{x}^i) \right ] \) and
\( \hat{\sigma}^2(z_1, z_2) = \frac{1}{|\mathcal{S}|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2 \). Both approximations are unbiased
only under the assumption that the points are uniformly distributed in
\([z_1, z_2]\). Under this assumption, and due to the law of large
numbers, the approximations become more robust, as the bins become
wider. On the other hand, bins should group local explanations with
similar effects and enlarging a bin is under the danger of adding
nuisance uncertainty. Therefore, we set-up the following optimization
problem:
\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of the cost of each bin. The cost of each bin is the
mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term
\(0.8 \leq \tau_K \leq 1\) acts as a tie-braker in cases of two
candidate bins with equal mean variance, favoring the selection of the
bigger bin. The constraint of at least \(N_{\mathtt{PPB}}\) points per
bin sets the lowest-limit for a \textit{robust} estimation.

@@

For achieving a computationally-grounded solution we set a threshold
\(K_{max}\) on the maximum number of bins which also discretizes the
solution space. The width of the bin can take discrete values that are
multiple of the minimum step
\(u = \frac{x_{s, max} - x_{s, min}}{K}\). For defining the solution,
we use two indexes. The index \(i \in \{0, \ldots, K_{max}\}\) denotes
the \(i\)-th limit \((z_i)\) and the index
\(j \in \{0, \ldots, K_{max}\} \) denotes the \(j\)-th position along
the axis \((x_j = x_{s,min} + j \cdot u)\). The recursive cost
function \(T(i,j)\) is the cost of setting the \(i\)-th limit \(z_i\)
to \(x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
\end{equation}
%
where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
with limits \([x_l, x_j)]\):

@@

% \begin{equation}
%   \label{eq:1}
%   \mathcal{L} = \sum_{k=0}^K \hat{\sigma}_k(z_{k-1}, z_k)
% \end{equation}
% %
% The constraints are that all bins must include more than \(\tau\)
% points, i.e., \(|\mathcal{S}_k| \geq \tau\).

% \(\mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k)\) is the sum
% of the uncertainty of the local explanations \(\mathcal{H}\) due to
% heterogeneous effects and the added nuisance uncertainty due to
% bin-splitting \(\mathcal{H}_n\). Given that \(\mathcal{H}\) is not
% affected by interval splitting, optimizing \(\mathcal{L}\) equals to
% finding the sequence of non-overlaping bins that add the minimum
% nuisance uncertainty. The restriction of (\ref{eq:opt}) secures that
% each bin is populated with enough samples for a robust estimation,
% which is set by the parameter \(N\). In other words, the user makes
% the following proposal; \textit{find the bin sequence that adds the
%   minimum nuisance uncertainty, given that I want at least \(N\)
%   points per bin}.

@@

\begin{itemize}
\item Computational complexity
\end{itemize}

\subsection{Visualization and Interpretation}
\label{sec:visualization}

\begin{itemize}
\item Discuss about the meaning of ALE, to find intervals with some effect
\end{itemize}

@@

In this simulation, we will compare NAME method against PDP-ICE in
quantifying the main effect \(f_\mu\) and the uncertainty
\(f_\sigma^2\). Since there is ambiguity about the ground-truth
first-order effect in cases of correlated
features, e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit
ourselves to the following piecewise linear function,

@@

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
zero otherwise. In each position, there is total absence of
heterogeneous effects; therefore, the uncertainty of the global
explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:ex-synth-1-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the regions with
constant-effect.

@@

\paragraph{Discussion.}

In purpose, the examples above do not cover the case where there is an
interaction term includes correlated features, i.e. \(x_1x_2\). In
this scenario, there is an open debate about the ground-truth
effect~\citep{Gromping2020MAEP}.

@@

\paragraph{Discussion.}

@@

we also
apply the methods on some real-world datasets. In these cases, it is
infeasible to obtain ground truth information, therefore, we perform
evaluation at two levels. First, we compare visually NAME versus PDP
with ICE plots. Second, we hold as ground-truth the effects computed
on the full training-set and we, then, perform subsampling to evaluate
the robustness of the bin-splitting method.

>>>>

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one the ALE approximation. Regarding the
definition ALE does not inform the user about the level of
heterogeneous effects hidden behind the average effect (global
explanation). In contrast, in the case of PDP, the heterogeneous
effects can be (partially) spotted by exploring the Individual
Conditional Expectations (ICE)\citep{goldstein2015peeking}. Regarding
the approximation, e.g.~the estimation of ALE from the limited samples
of the training set, ALE requires an important additional step, called
\textit{bin-splitting}. Bin-splitting consists of partitioning the
axis of the feature of interest in a sequence of non-overlapping
intervals and estimating a single effect in each one, from the
population of samples that fall inside. Specifying an approriate
sequence of intervals is of particular importance, since ALE's
interpretation is meaningful only inside each interval
\citep{molnar2022}. However, this crucial step has not raised the
appropriate attention, making the approximation vulnerable to
potential misinterpretations.

@@

\paragraph{Notation.} We refer to random variables (rv) using
uppercase \( X \), to simple variables with plain lowercase \( x \)
and to vectors with bold \( \xb \). Often, we partition the input
vector \(\xb \in \mathbb{R}^D\) to the feature of interest
\(x_s \in \mathbb{R} \) and the rest of the features
\(\xc \in \mathbb{R}^{D-1}\). For convenience we denote it as
\((x_s, \mathbf{x}_c)\), but we clarify that it implies the vector
\((x_1, \cdots , x_s, \cdots, x_D)\). Equivalently, we denote the
corresponding rv as \(\mathbf{X} = (X_s, \mathbf{X}_c)\). When we
refer to an instance of the training set, we use
\(\xb^i= (\xc^i, x_s^i) \). The black-box function is
\(f : \mathbb{R}^D \rightarrow \mathbb{R}\) and the FE of the \(s\)-th
feature is \(f^{\mathtt{<method>}}(x_s)\), with \(\mathtt{<method>}\)
indicating the particular method in use.\footnote{An extensive list of all
  symbols used in the paper is provided at the Appendix.}

\subsection{Feature Effect Methods And ALE Definition}
\label{sec:feat-effect-meth}

The three well-known feature effect methods are: PDP, MP and ALE. PDP
formulates the FE of the \(s\)-th attribute as an expectation over the
marginal distribution \(\mathbf{X}_c\), i.e.,
\(f^{\mathtt{PDP}}(x_s) =
\mathbb{E}_{\mathbf{X}_c}[f(x_s,\mathbf{X}_c)]\), whereas MP
formulates it as an expectation over the conditional
\(\mathbf{X}_c|X_s\), i.e.,
\(f^{\mathtt{MP}}(x_s) = \mathbb{E}_{\mathbf{X}_c|x_s}[f(x_s,
\mathbf{X}_c)]\). ALE defines the local effect of the \(s\)-th feature
at point \(x_s = z\) as \(f^s(z, \xc) = \frac{\partial f}{\partial x_s} (z, \xc)\). All the
local explanations at \(z\) are, then, weighted by the conditional
distribution \(p(\xc|x_s=z)\) and are averaged, to produce the
averaged effect \(\mu(z)\). Finally, ALE is the accumulation of the
averaged local effects:

@@

FE methods pose the question \textit{what is average (global) effect
  on the output, if the value of a specific feature is
  increased/decreased}. After that, it comes naturally to also ask
about \textit{to what extent the individual (local) effects deviate
  from the average}. For this reason, the quantification of the
uncertainty has attracted a lot of interest.

ICE and d-ICE\citep{goldstein2015peeking} provide a set of curves that
are plot on top-of the PDP. Both methods produce one curve for each
instance of the training set;
\(f^{\mathtt{ICE}}_i(x_s) = f(x_s, \xc^i)\) for ICE and
\(f^{\mathtt{d-ICE}}_i(x_s) = \frac{\partial f}{\partial x_s} (x_s,
\xc^i)\) for d-ICE. The user then visually observes if the (d)-ICE
curves are homogeneous and to what extent they deviate from the
PDP. Some methods try to automate the aforementioned visual
exploration, by grouping (d-)ICE into clusters~\citep{molnar2020model,
  herbinger2022repid, britton2019vine}. Some other approaches, like
H-Statistic\citep{friedman2008predictive}, Greenwell's interaction
index\citep{greenwell2018simple} or SHAP interaction
values\citep{lundberg2018consistent}, quantify the level of
interaction between the input features, with an interaction value. A
strong interaction index is an indicator for the existence of
heterogeneous effects.

The aforementioned approaches are under two limitations; They either
do not quantify the uncertainty of the FE directly or they are based
on PDPs, and, therefore, they are subject to the failure modes of PDPs
in cases of correlated features\citep{baniecki2021fooling}. To the
best of our knowledge, no work exist so far that quantifies the
heterogeneous effects based on the formulation of ALE.

@@

In real ML scenarios, the FE is estimated from the limited instances
of the training set. Therefore, \citep{apley2020visualizing} proposed
dividing the \(s\)-th axis in \(K\) bins and estimating the local
effects in each bin by evaluating the black box-function at the bin
limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards, (cite) proposed the Differential ALE (DALE)
approximation for computing the local effects on the training
instances using auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

Both approximations ask from the user to blindly decide the number of
bins \(K\) for splitting the axis into equally-sized bins, an approach
with crucial limitations. Firstly, it is vulnerable to poor bin-effect
approximations that lead to a misleading ALE plot. Setting \(K\) to a
small value may hide fine-grain effects due to large bins and setting
\(K\) to a high value leads to estimations of many bin-effects from
very limited samples. In general, the user may face contradictory
explanations for differnt \(K\) without a clue to decide which one to
trust. Secondly, as indicated by~\citep{molnar2022}, the ALE's feature
effect can only be interpreted in local regions. Therefore, for better
interpretation, the extend of these regions should be determined
in a data-driven way.

@@

The NAME method extends ALE for quantifying the level of heterogeneous
effects (uncertainty) and automating the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define NAME,
presenting the component that extends ALE with uncertainty
quantification. In Section~\ref{sec:interval-based-estimation}, we
provide an alternative definition of NAME based on variable-size bins,
providing important remarks and proofs for connecting it with the
initial definition. In Section~\ref{sec:interval-spliting}, we define
the problem of optimal bin-splitting and we provide an algorithmic
solution to the problem.

@@

ALE does not shed light to potential heterogeneous effects that
contribute to the average explanation. Therefore, we extend ALE
definition of Eq.~(\ref{eq:ALE}) with a component that quantifies the
level of heterogeneous effects. We denote \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
NAME method formulates the effect at a specific point \(x_s\) with a
compact entity that consists of the average effect and the uncertainty
\(\mu(z) \pm \sigma(z)\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{METHOD}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

@@

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the interval \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). We define the mean effect (bin-effect) and the
uncertainty (bin-uncertainty) inside an interval \([z_1, z_2)\) as:

@@

%
We also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points that
partition the \(s\)-th feature axis into a sequence \(K\)
variable-size intervals, i.e. \(\mathcal{Z}_s = \{z_0, z_K\}\), where
\(z_0=x_{s,min}\) and \(z_K=x_{s,max}\). Based on the above, we
redefine NAME using an interval-based formulation
\(\tilde{f}^{\mathtt{METHOD}}(x_s):=
\tilde{f}^{\mathtt{ALE}}_{\mu}(x_s) \pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\textit{If we define the residual \(\rho(z)\) as the
difference between the expected effect at \(x_s\) and the mean
expected effect at the interval, i.e
\(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the accumulated variance
\(\sigma^2(z_1, z_2)\) equals to the accumulation of the point-wise
variances plus the squared residuals inside the interval:
\begin{equation}
 \sigma^2(z_1, z_2) = \int_{z_1}^{z_2} \sigma^2(z) + \rho^2(z) \partial z
\end{equation}}
%

@@

\textit{If we define as \(\rho^2(z_{k-1}, z_{k})\) the accumulated residual of
a bin, i.e.
\(= \rho^2(z_{k-1}, z_{k}) = \int_{z_{k-1}}^{z_k} \rho^2(z)\partial
z\), then the two definitions of NAME are equivalent, i.e.,
\(f^{\mathtt{NAME}} = \tilde{f}^{\mathtt{NAME}}\), if and only if the
sequence of limits \(\mathcal{Z}\) is such that
\(\sum_{k=1}^K \rho^2(z_{k-1}, z_{k}) = 0\).}

@@

In this section we formulate bin-splitting as an unsupervised
clustering problem. The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}_k\) of the dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \}
\):

\begin{equation}
  \label{eq:2}
  \hat{\mu}(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
  \sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left [ \dfdx(\mathbf{x}^i)
  \right ]
\end{equation}
%
\begin{equation}
  \label{eq:3}
  \hat{\sigma}^2(z_{k-1}, z_k) = \frac{1}{|\mathcal{S}_k|}
\sum_{i:\mathbf{x}^i \in \mathcal{S}_k} \left ( \dfdx(\mathbf{x}^i) -
  \hat{\mu}(z_1, z_2) \right )^2
\end{equation}

Both approximations are unbiased only under the assumption that the
points are uniformly distributed inside the intervals. Under this
assumption, and due to the law of large numbers, the approximations
become more robust, as the bins become wider. On the other hand, bins
should group local explanations with similar effects and enlarging a
bin is under the danger of adding nuisance uncertainty. Therefore, we
set-up the following optimization problem:
\begin{equation}
  \label{eq:opt}
\begin{aligned}
  \min_{ \mathcal{Z} = \{z_0, \ldots, z_K\}} \quad & \mathcal{L} = \sum_{k=1}^K \hat{\sigma}^2(z_{k-1}, z_k) \Delta z_k \\
  \textrm{where} \quad & \Delta z_k = z_k - z_{k-1} \\
  & \tau_k = 1 - 0.2 \frac{|S_k|}{N} \\
  \textrm{s.t.} \quad & |\mathcal{S}_k| \geq N_{\mathtt{NPB}}\\
                                     & z_0 = x_{s,min}\\
                                     & z_K = x_{s, max}
\end{aligned}
\end{equation}
%
We search for the sequence of intervals \(z_0, \ldots, z_K\) that
minimizes the sum of the cost of each bin. The cost of each bin is the
mean variance \(\sigma^2_k\) scaled by the bin length \(\Delta z_k\)
and discounted by the term \(\tau_k\). The term \(\tau_K\) acts as a
tie-braker in cases of two candidate bins with equal mean variance,
favoring the selection of the bigger bin. The constraint of at least
\(N_{\mathtt{PPB}}\) points per bin sets the lowest-limit for a
\textit{robust} estimation.

@@

For achieving a computationally-grounded solution we set a threshold
\(K_{max}\) on the maximum number of bins which also discretizes the
solution space. The width of the bin can take discrete values that are
multiple of the minimum step
\(u = \frac{x_{s, max} - x_{s, min}}{K_{max}}\). For defining the
solution, we use two indexes. The index
\(i \in \{0, \ldots, K_{max}\}\) denotes the point \((z_i)\) and the
index \(j \in \{0, \ldots, K_{max}\} \) denotes the position of the
\(j\)-th multiple of the minimum step, i.e., 
\(x_j = x_{s,min} + j \cdot u\). The recursive cost function
\(T(i,j)\) is the cost of setting \(z_i=x_j\):
\begin{equation}
  \label{eq:recursive_cost}
  \mathcal{T}(i,j) = \mathrm{min}_{l \in \{0, \ldots, K_{max}\}} \left [ \mathcal{T}(i-1, l) + \mathcal{B}(x_l, x_j) \right ]
\end{equation}
%
where \(\mathcal{T}(0,j)\) equals zero if \(j=0\) and \(\infty\) in
any other case. \(\mathcal{B}(x_l, x_j)\) denotes the cost of creating a bin
with limits \([x_l, x_j)]\):

@@

@@

\begin{itemize}
\item Discuss more aspects (e.g. Computational complexity)
\end{itemize}

@@

In this simulation, we will compare NAME method against PDP-ICE in
quantifying the main effect and the uncertainty. Since there is
ambiguity about the ground-truth first-order effect in cases of
correlated features,
e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
to the following piecewise linear function,

@@

We test the feature effect when no interaction term is apparent,
i.e. \(\alpha=0\), and the weights are equal \(a_1=a_2=1\).  In this
case, the ground truth effect \(f_\mu^{\mathtt{GT}}(x_1)\) is: \(x_1\)
when \(0 \leq x_1 < 0.25\), \(-x_1\) when \(0.25 \leq x_1 < 0.5\) and
zero otherwise. In each position, there is total absence of
heterogeneous effects; therefore, the uncertainty of the global
explanation is zero \(f^{\mathcal{GT}}_{\sigma^2}(x_1) = 0\). In
Figure~\ref{fig:synth-ex-1-case-1}, we observe that PDP main effect is
wrong and ICE plots imply the existence of heterogeneous effects. In
contrast, ALE captures correctly the average effect and the zero
uncertainty and it also groups perfectly the regions with
constant-effect.

@@

% \paragraph{Discussion.}

% In purpose, the examples above do not cover the case where there is an
% interaction term includes correlated features, i.e. \(x_1x_2\). In
% this scenario, there is an open debate about the ground-truth
% effect~\citep{Gromping2020MAEP}.

@@

@@

In Real-World Datasets it is infeasible to obtain ground truth
information, therefore, we perform evaluation at two levels. First, we
compare visually NAME versus PDP with ICE plots. Second, we hold as
ground-truth the effects computed on the full training-set and we,
then, perform subsampling to evaluate the robustness of the
bin-splitting method.
